---
title: 'Superfund Sites: Hennepin, Ramsey, Washington'
author: "Erin Franke"
date: "2023-04-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and Data

```{r}
library(tidyverse)
library(sf)
library(USAboundaries)
```

```{r, eval=FALSE}
threeM <- read_csv("../data/superfund_site_data/3m_chemolite.csv") %>%
  mutate(site = "3M Chemolite", county = "Washington", SYS_LOC_CODE = as.numeric(SYS_LOC_CODE))
oakdale <- read_csv("../data/superfund_site_data/oakdale.csv") %>%
  mutate(site = "3M Oakland", county = "Washington")
ashland <- read_csv("../data/superfund_site_data/ashland.csv") %>%
  mutate(site = "Ashland Oil - Park Penta", county = "Washington")
bayport <- read_csv("../data/superfund_site_data/bayport.csv") %>%
  mutate(site = "Baytown Township", county = "Washington")
lakeland <- read_csv("../data/superfund_site_data/lakeland.csv") %>%
  mutate(site = "Lakeland", county = "Washington")

lyndale55 <- read_csv("../data/superfund_site_data/66th.csv") %>%
  mutate(site = "55th St & Lyndale Ave S", county = "Hennepin")
vincent66 <- read_csv("../data/superfund_site_data/lynndale55.csv") %>%
  mutate(site = "66th St. & Vincent Ave", county = "Hennepin")
cedarServices <- read_csv("../data/superfund_site_data/cedarservices.csv") %>%
  mutate(site = "Cedar Services (MDA)", county = "Hennepin")
chemMarketing <- read_csv("../data/superfund_site_data/chemMarketing.csv") %>%
  mutate(site = "Chemical Marketing Corp of America", county = "Hennepin")
cmcHeartland <- read_csv("../data/superfund_site_data/cmcHeartland.csv") %>%
  mutate(site = "CMC Heartland Lite Yard (MDA)", county = "Hennepin")
generalMills <- read_csv("../data/superfund_site_data/generalMills.csv") %>%
  mutate(site = "General Mills/Henkel Corp. site", county = "Hennepin")
goldEagleH <- read_csv("../data/superfund_site_data/goldenEagleH.csv") %>%
  mutate(site = "Gold Eagle Cleaners", county = "Hennepin")
hmong <- read_csv("../data/superfund_site_data/hmong.csv") %>%
  mutate(site = "Hmong Shopping Center/Pilgrim Cleaners", county = "Hennepin")
minnegasco <-  read_csv("../data/superfund_site_data/minnegasco.csv") %>%
  mutate(site = "Minnegasco", county = "Hennepin")
pilgrim <- read_csv("../data/superfund_site_data/pilgrim.csv") %>%
  mutate(site = "Pilgrim Cleaner's site", county = "Hennepin")
precision <- read_csv("../data/superfund_site_data/precision.csv") %>%
  mutate(site = "Precision Plating Inc.", county = "Hennepin")
reilly <- read_csv("../data/superfund_site_data/reilly.csv") %>%
  mutate(site = "Reilly Tar", county = "Hennepin")
schloff <- read_csv("../data/superfund_site_data/schloff.csv") %>%
  mutate(site = "Schloff Chemical", county = "Hennepin")
sehennepin <- read_csv("../data/superfund_site_data/seHennepin.csv") %>%
  mutate(site = "Southeast Hennepin Area groundwater and vapor site", county = "Hennepin")
springpark <- read_csv("../data/superfund_site_data/springpark.csv") %>%
  mutate(site = "Spring Park Municipal Wells", county = "Hennepin")
saintlouis <- read_csv("../data/superfund_site_data/saintlouis.csv") %>%
  mutate(site = "St. Louis Park solvent plume", county = "Hennepin", SYS_LOC_CODE = as.numeric(SYS_LOC_CODE))
superior <- read_csv("../data/superfund_site_data/superior.csv") %>%
  mutate(site = "Superior Plating Inc.", county = "Hennepin")
tonkaMain <- read_csv("../data/superfund_site_data/tonkaMain.csv") %>%
  mutate(site = "Tonka Main Plant", county = "Hennepin")
universalPlating <- read_csv("../data/superfund_site_data/universal.csv") %>%
  mutate(site = "Universal Plating", county = "Hennepin")
whiteway <- read_csv("../data/superfund_site_data/whiteway.csv") %>%
  mutate(site = "White Way Cleaners", county = "Hennepin")

arcadeHawthrone <- read_csv("../data/superfund_site_data/arcade.csv") %>%
  mutate(site = "Arcade & Hawthrone Ave E", county = "Ramsey")
bellLumber <- read_csv("../data/superfund_site_data/bell_lumber.csv") %>%
  mutate(site = "Bell Lumber & Pole Company", county = "Ramsey")
centerville <- read_csv("../data/superfund_site_data/centerville.csv") %>%
  mutate(site = "Centerville Road Dump", county = "Ramsey")
fishHatchery <- read_csv("../data/superfund_site_data/fish_hatchery.csv") %>%
  mutate(site = "Fish Hatchery Dump", county = "Ramsey")
highway96 <- read_csv("../data/superfund_site_data/highway96.csv") %>%
  mutate(site = "Highway 96 Dump", county = "Ramsey")
macGillis <- read_csv("../data/superfund_site_data/macGillis.csv") %>%
  mutate(site = "MacGillis and Gibbs waste site", county = "Ramsey")
pigseye <- read_csv("../data/superfund_site_data/pigseye.csv") %>%
  mutate(site = "Pig's Eye Landfill", county = "Ramsey")
tcaap <- read_csv("../data/superfund_site_data/tcaap.csv") %>%
  mutate(site = "Twin Cities Army Ammunition Plant (TCAAP)", county = "Ramsey")
universityPascal <- read_csv("../data/superfund_site_data/universityPascal.csv") %>%
  mutate(site = "University Ave & Pascal St.", county = "Ramsey")

superfund <- bind_rows(arcadeHawthrone, ashland, bayport, bellLumber, cedarServices, centerville, chemMarketing, cmcHeartland, fishHatchery, generalMills, goldEagleH, highway96, hmong, lakeland, lyndale55, macGillis, minnegasco, oakdale, pigseye, pilgrim, precision, reilly, saintlouis, schloff, sehennepin, springpark, superior, tcaap, threeM, tonkaMain, universalPlating, universityPascal, vincent66, whiteway) %>%
   mutate(commonName = case_when(ANALYTE_NAME == "Perfluorobutanoic acid (PFBA)" ~ "PFBA", 
                                ANALYTE_NAME == "Perfluorooctanoic acid (PFOA)" ~ "PFOA", 
                                ANALYTE_NAME == "Perfluoropentanoic acid (PFPeA)" ~ "PFPeA",
                                ANALYTE_NAME == "Perfluorohexanoic acid (PFHxA)" ~ "PFHxA", 
                                ANALYTE_NAME == "Perfluorooctanesulfonate (PFOS)" ~ "PFOS", 
                                ANALYTE_NAME == "Perfluorohexanesulfonate (PFHxS)" ~ "PFHxS", 
                                ANALYTE_NAME == "Perfluorobutanesulfonate (PFBS)" ~ "PFBS"), 
          RESULT_NUMERIC = case_when(RESULT_UNIT == "ng/L" ~ RESULT_NUMERIC/ 1000, 
                                    TRUE ~ RESULT_NUMERIC))

save(superfund, file = "../data/superfund_site_data/superfund.rds")
```

Brooklyn Park Dump: Hennepin
Edina Well Field site: Hennepin
Gold Eagle Cleaners: Ramsey
Honeywell Inc: Hennepin
Hospital Linen: Ramsey
Joslyn Mfg. and Supply Co. OU1: Hennepin
Lyndale Avenue Corridor: Hennepin
MIBCO site: Hennepin
Minnetonka Blvd & Raleigh Ave: Hennepin
Pure Oil Bulk Facility: Hennepin
St. Paul Levee property: Ramsey
Valentine Clark Corp: Ramsey

Load in Superfund Data

```{r}
load("../data/superfund_site_data/superfund.rds")
```

```{r}
counties <- us_counties(resolution = "high", states = c("Minnesota")) %>%
  st_transform(crs = 6783)
counties_cropped <- st_crop(counties, xmin = -93.7, xmax=-92.6, ymin = 44.7, ymax=45.2)

superfund_loc <- superfund %>%
  filter(!is.na(LATITUDE), !is.na(LONGITUDE), LATITUDE > 44, LATITUDE < 46, LONGITUDE < -92, LONGITUDE > -94)
superfund_loc_pfas <- superfund_loc %>%
  filter(ANALYTE_NAME %in% c("Perfluorohexanoic acid (PFHxA)", "Perfluorohexanesulfonate (PFHxS)", "Perfluorooctanesulfonate (PFOS)", "Perfluorooctanoic acid (PFOA)", "Perfluoropentanoic acid (PFPeA)", "Perfluorobutanesulfonate (PFBS)", "Perfluorobutanoic acid (PFBA)"))

superfund_loc <- st_as_sf(superfund_loc, coords = c("LONGITUDE", "LATITUDE"), crs = 6783)
superfund_loc_pfas <- st_as_sf(superfund_loc_pfas, coords = c("LONGITUDE", "LATITUDE"), crs = 6783)
```

## Plot superfund sites 

Those that have been sampled from

```{r}
ggplot()+
  geom_sf(data = counties_cropped, color = "navajowhite", fill = "ivory", size = 0.5)+
  geom_sf(data = superfund_loc, size= 0.9, aes(color = site), alpha = 0.5)+
  theme(legend.position = "none", 
        axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())+
  labs(title = "Samples from superfund sites in Washington, Hennepin, and Ramsey Counties")
```

Superfund sites that have had pfas samples

```{r}
ggplot()+
  geom_sf(data = counties_cropped, color = "navajowhite", fill = "ivory", size = 0.5)+
  geom_sf(data = superfund_loc_pfas, size= 0.9, aes(color = site), alpha = 0.5)+
  theme(legend.position = "none", 
        axis.line = element_blank(), 
        axis.text = element_blank(), 
        axis.ticks = element_blank())+
  labs(title = "PFAS samples from superfund sites", subtitle = "Sites located in Hennepin, Washington, and Ramsey counties")
```

## Look at pfas in superfund sites in Washington County

```{r}
washington <- superfund_loc_pfas %>% filter(county == "Washington")

counties_cropped <- st_crop(counties, xmin = -93.2, xmax=-92.7, ymin = 44.7, ymax=45.35)

ggplot()+
  geom_sf(data = counties_cropped, color = "navajowhite", fill = "ivory", size = 0.5)+
  geom_sf(data = washington, size= 0.9, aes(color = site), alpha = 0.5)+
  labs(title = "PFAS from superfund sites in Washington County")+
  theme(axis.line = element_blank(), 
        axis.ticks = element_blank(),
        axis.text = element_blank())
```
